---
title: "Calculation of minimum inhibitory concentrations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculation of minimum inhibitory concentrations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Aim
Jacobo de la Cuesta-Zuluaga. March 2023.

This notebook introduces the calculation of minimum inhibitory concentrations. 
from bacterial growth curves. This functionality is an extension to the basic
loading and cleaning of bacterial growth curves and the calculation of areas under
the curve introduced in the `Bacterial_curve_Analysis` vignette. For a detailed
explanation of these steps, please refer to this file.

```{r}
# Load R packages needed in this vignette
library(neckaR)
library(tidyverse)
library(conflicted)
```

```{r}
# Solve conflicts
conflict_prefer("filter", "dplyr")
```

# Load data
```{r}
# List file locations
extdata = system.file("extdata", package = "neckaR")
data_dir = file.path(extdata, "MICs")

# List raw plate reader files
Raw_Data_files = list.files(data_dir, pattern = "Run")

# Tab1
Runs = file.path(data_dir, "Tab1.xlsx")

# Tab2
Layout = file.path(data_dir, "Tab2.xlsx")
```

```{r}
# Create master data frame
MIC_master_df = Make_master_df(Data_folder = data_dir, 
															 Data_files = Raw_Data_files, 
															 Runs_path = Runs, 
															 Layout_path = Layout, 
															 Design_tab_col = "Design")

```
```{r}
MIC_master_df %>% 
	head()
```
# Plot raw curves
```{r}
# Plot raw curves
MIC_master_plots = Make_curve_plots(MIC_master_df, 
																		save_plots = FALSE,
																		replicate_variable = "Replicate")

# Show just the first 3 graphs
MIC_master_plots[1:3]
```

# Adjust ODs
```{r}
MIC_adjusted_df = Adjust_OD(curves_df = MIC_master_df, 
														control_factor = "Treatment", 
														control_level = "Water", 
														offset_control = 0.02) %>% 
	Calculate_lag(offset_control = 0.02)

# Print head
MIC_adjusted_df %>% 
	head()
```

# Plot cut-offs 
```{r}
cutoff_plot = Make_cutoff_plots(curves_df = MIC_adjusted_df, 
																save_plots = FALSE,  
																vline = "cutoff")

cutoff_plot
```

# Adjust cut-off values
```{r}
# Create list
replace_values = list(c(210, 10),
											c(201, 7),
											c(209, 6),
											c(701, 10),
											c(403, 10))

# replace values
MIC_readjusted_df = Fix_cutoff(MIC_adjusted_df, 
															 replace_values)
```

## Plot curves up to cut-off times
```{r}
# cut all curves at the cut-off
# Note that this is used only for plotting
MIC_cutoff_df = MIC_adjusted_df %>% 
  filter(Time < (cutoff_time + 0.5)) %>%
  arrange(Time)

# Plot
MIC_cutoff_df %>% 
  filter(Control == TRUE) %>% 
  Make_cutoff_plots(save_plots = FALSE,  
                    plot_name = "Truncated_control_cutoff")
```
# Growth curve quality control
## Mark artefacts

```{r}
# Run mark artefacts function
MIC_marked_df = Mark_artefacts(MIC_readjusted_df,
															 sum_inc = 3,
															 increased_sd = 2,
															 t0_sd = 3,
															 p_delta = 1e-3,
															 p_delta2 = 1e-3)
```

```{r}
# Plot marked curves
MIC_filtered_plots = Make_filtered_plot(MIC_marked_df,
																				save_plots = FALSE,
																				replicate_variable = "Replicate")

MIC_filtered_plots[1:3]

```

# Calculate AUC
```{r}
MIC_filtered_df = MIC_marked_df %>% 
  filter(discard_conc == FALSE)
```

```{r}
MIC_AUC_df = Calculate_AUC(MIC_filtered_df)
```

# Detect MIC
```{r}
# Create df with max OD and cut-off times of each curve
MIC_max_OD_df = MIC_filtered_df %>% 
	group_by(RRPPRCC) %>% 
	slice(which.min(cutoff_time - Time)) %>% 
	ungroup() %>% 
	select(RRPPRCC, Strain, Treatment, Concentration, Control, Replicate, 
				 Max_OD = OD, Max_ODc01 = ODc01, cutoff_time, lag_time)  

# Select only normalized AUC
MIC_selected_AUC_df = MIC_AUC_df %>% 
	select(RRPPRCC, normAUC)

# Combined df
MIC_AUC_final_df = left_join(MIC_max_OD_df, MIC_selected_AUC_df, by = "RRPPRCC")

MIC_AUC_final_df %>% 
	head()
```

```{r}
MIC_df  = Detect_MIC(curves_df = MIC_AUC_final_df, 
										 threshold = 0.75, 
										 concentration_var = Concentration, 
										 group_var = c("Treatment", "Strain"), 
										 control_var = Control)
```

```{r}
MIC_df %>% 
	arrange(Strain)
```

