---
title: "Calculation of minimum inhibitory concentrations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculation of minimum inhibitory concentrations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Aim
Jacobo de la Cuesta-Zuluaga. March 2023.

This notebook introduces the calculation of minimum inhibitory concentrations. 
from bacterial growth curves. This functionality is an extension to the basic
loading and cleaning of bacterial growth curves and the calculation of areas under
the curve introduced in the `Bacterial_curve_Analysis` vignette. For a detailed
explanation of these steps, please refer to this file.

# Libraries

```{r}
# Load R packages needed in this vignette
library(neckaR)
library(tidyverse)
library(conflicted)
```

```{r}
# Solve conflicts
conflict_prefer("filter", "dplyr")
```

# Load data
First, we will load the data. Recall that we need to specify the paths of 
the *Runs* a.k.a `Tab1` and *Layout* a.k.a `Tab2` files, in addition to
the raw files coming from the plate reader.

In this case, the *Layout* table should contain data regarding the concentration
of the treatments used.
```{r}
# List file locations
extdata = system.file("extdata", package = "neckaR")
data_dir = file.path(extdata, "MICs")

# List raw plate reader files
Raw_Data_files = list.files(data_dir, pattern = "Run")

# Tab1
Runs = file.path(data_dir, "Tab1.xlsx")

# Tab2
Layout = file.path(data_dir, "Tab2.xlsx")
```


We create the master data frame
```{r}
# Create master data frame
MIC_master_df = Make_master_df(Data_folder = data_dir, 
															 Data_files = Raw_Data_files, 
															 Runs_path = Runs, 
															 Layout_path = Layout, 
															 Design_tab_col = "Design")

```
```{r}
MIC_master_df %>% 
	head()
```
# Plot raw curves
We then plot and inspect the raw curves
```{r}
# Plot raw curves
MIC_master_plots = Make_curve_plots(MIC_master_df, 
																		save_plots = FALSE,
																		replicate_variable = "Replicate")

# Show just the first 3 graphs
MIC_master_plots[1:3]
```

# Adjust ODs
Next, we will adjust the optical density (OD). For this, we need to specify the
name of the column containing the treatment information and which of them corresponds
to the control.
```{r}
MIC_adjusted_df = Adjust_OD(curves_df = MIC_master_df, 
														control_factor = "Treatment", 
														control_level = "Water", 
														offset_control = 0.02) 

# Print head
MIC_adjusted_df %>% 
	head()
```

# End of exponential phase
Since areas under the curve (AUC) are calculated based on the end of the exponential
phase of the controls, we need to plot and inspect the cut of values.
```{r}
cutoff_plot = Make_cutoff_plots(curves_df = MIC_adjusted_df, 
																save_plots = FALSE,  
																vline = "cutoff")

cutoff_plot
```

## Adjust cut-off values
We can manually adjust the cut-off values in case we're not satisfied with 
the ones automatically detected by the functions.
```{r}
# Create list of values to replace
replace_values = list(c(210, 10),
											c(201, 7),
											c(209, 6),
											c(701, 10),
											c(403, 10))

# replace values
MIC_readjusted_df = Fix_cutoff(MIC_adjusted_df, 
															 replace_values)
```

## Plot curves up to cut-off times
```{r}
# cut all curves at the cut-off
# Note that this is used only for plotting
MIC_cutoff_df = MIC_adjusted_df %>% 
  filter(Time < (cutoff_time + 0.5)) %>%
  arrange(Time)

# Plot
MIC_cutoff_df %>% 
  filter(Control == TRUE) %>% 
  Make_cutoff_plots(save_plots = FALSE,  
                    plot_name = "Truncated_control_cutoff")
```
# Growth curve quality control
Once the cut-off values are set, we proceed to do the quality control of the 
growth curves.
```{r}
# Run mark artefacts function
MIC_marked_df = Mark_artefacts(MIC_readjusted_df,
															 sum_inc = 3,
															 increased_sd = 2,
															 t0_sd = 3,
															 p_delta = 1e-3,
															 p_delta2 = 1e-3)
```
## Plot marked curves
Recall that we can visually inspect the curves marked as abnormal. For methods
to override the identified curves, see the `Bacterial_curve_Analysis` vignette.
```{r}
# Plot marked curves
MIC_filtered_plots = Make_filtered_plot(MIC_marked_df,
																				save_plots = FALSE,
																				replicate_variable = "Replicate")

MIC_filtered_plots[1:3]

```

## Remove abnormal curves
```{r}
MIC_filtered_df = MIC_marked_df %>% 
  filter(discard_conc == FALSE)
```

# Calculate AUC
Once abnormal curves have been discarded, we can calculate the AUCs
```{r}
MIC_AUC_df = Calculate_AUC(MIC_filtered_df)
```

```{r}
MIC_AUC_df %>% 
	head()
```


We can create a data frame that contains information about the AUC, maximum OD and 
cut-off time for each individual curve. This table can be useful in downstream analyses.
```{r}
# Create df with max OD and cut-off times of each curve
MIC_max_OD_df = MIC_filtered_df %>% 
	group_by(RRPPRCC) %>% 
	slice(which.min(cutoff_time - Time)) %>% 
	ungroup() %>% 
	select(RRPPRCC, Strain, Treatment, Concentration, Control, Replicate, 
				 Max_OD = OD, Max_ODc01 = ODc01, cutoff_time)  

# Select only normalized AUC
MIC_selected_AUC_df = MIC_AUC_df %>% 
	select(RRPPRCC, normAUC)

# Combined df
MIC_AUC_final_df = left_join(MIC_max_OD_df, MIC_selected_AUC_df, by = "RRPPRCC")

MIC_AUC_final_df %>% 
	head()
```

# Detect MIC
Finally, we can use either the summary we just created or the table with
AUCs generated to determine the minimum inhibitory concentration for each
treatment
```{r}
MIC_df  = Detect_MIC(curves_df = MIC_AUC_final_df, 
										 threshold = 0.75, 
										 concentration_var = Concentration, 
										 group_var = c("Treatment", "Strain"), 
										 control_var = Control)
```

```{r}
MIC_df %>% 
	arrange(Strain)
```


# Session Info
```{r}
sessionInfo()
```

